Incorporating spatial distribution at age improves predictive ability of SDMs

### 2024-12-14 paper.qmd
update

As a first validation, we leveraged the natural recursive element of our data to quantify the ability of our models to predict the future distribution of ages. For each of the species - age models constructed above, we made predictions to the survey locations in the following year (e.g. age 3 hake in years 2007 -- 2018 used to predict the distribution of age 4 hake in 2008 -- 2019). We related the predicted probabilities of occurrence $\hat{p}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{p}_{\boldsymbol{s},t}$. We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$. The exponentiated slope parameter $exp(\beta_{1})$ represents a the change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival.

As a second validation, we compared aggregated predictions across all ages to a SDM model that used total biomass as a response.

### Links to fisheries data?

@@

Estimates of age composition by year may be useful in tracking the distribution of cohorts through time, as well as examining variability in distributions of particular age classes over time. For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, @fig-hake-spatial-composition) and strong sablefish cohorts (2013, 2016, 2020, @fig-sablefish-spatial-composition). \[expand on this\]

When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, we see a decrease in the spatial and spatiotemporal variances as well as a decrease in the spatial range (@fig-spatial-parameters). Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, suggesting that the fields estimated for older fish are less variable. Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ \~ 0.13). The decline in range with age implies that spatial distributions also become more concentrated as a function of age (@fig-spatial-parameters).

Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish (@fig-glm-coefficients).

@@

\break

```{r label="fig-hake-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for Pacific hake ages 1 -- 4.", echo=FALSE}
knitr::include_graphics("plots/hake_spatial_anomalies.png")
```

\break

```{r label="fig-sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for sablefish ages 0 -- 5.", echo=FALSE}
knitr::include_graphics("plots/sablefish_spatial_anomalies.png")
```

\break

```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial composition for Pacific hake; for plotting purposes only ages 1 - 5 and the last 10 years of available predictions are shown.", echo=FALSE}
knitr::include_graphics("plots/spatial_composition_Pacific hake.png")
```

\break

```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial composition for sablefish; for plotting purposes only ages 0 - 4 and the last 10 years of available predictions are shown.", echo=FALSE}
knitr::include_graphics("plots/spatial_composition_sablefish.png")
```

\break

@@

\break

@@

\break

@@

>>>>

As a first validation, we leveraged the natural recursive element of our data to quantify the ability of our models to predict the future distribution of ages. For each of the species - age models constructed above, we made predictions to the survey locations in the following year (e.g. age 3 hake in years 2007 -- 2018 used to predict the distribution of age 4 hake in 2008 -- 2019). We related the predictions $\hat{p}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{p}_{\boldsymbol{s},t}$. We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$. The exponentiated slope parameter $exp(\beta_{1})$ represents a change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival. As a diagnostic check and to understand whether there was a potential time component, we fit a second model as a GLMM with R package glmmTMB; this second model was identical to our Poisson GLM with random effects in $\beta_{0}$ and $\beta_{1}$, where year was used as the grouping variable.

### Using predictions to quantify bycatch risk

To illustrate potential benefits of this kind of modeling, we used the above approach to generate spatiotemporal predictions of age-1 sablefish (from age-0 sablefish the year prior). \[Need some text here on background why sablefish as bycatch is important\]. Gridded predictions from the Poisson GLM represent the expected number of fish per unit of effort in each location. To turn these into a more useful measure of risk, we calculate the total number of age-1 sablefish per unit of effort within a given radius of major fishing ports on the US west coast. For consistency with previous work \[TODO: cite ESR\] we used a radius of 232km.

@@

Estimates of age composition by year may be useful in tracking the distribution of cohorts through time, as well as examining variability in distributions of particular age classes over time. For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, @fig-hake-spatial-composition) and strong sablefish cohorts (2013, 2016, 2020, @fig-sablefish-spatial-composition). These predictions are also useful in revealing subtle differences between strong cohort years. For example, the 2009 cohort of Pacific hake (age 1 in 2010) were predominantly distributed in the northern part of the domain, while the 2013 cohort had hotspots along the entire US west coast. Similarly for sablefish, the 2016 cohort appeared to be more concentrated in the north, versus the more widely distributed 2013 cohort.

When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range (@fig-spatial-parameters). Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, meaning that spatial fields estimated for older fish are less variable. Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ \~ 0.13). The decline in range with age implies that spatial distributions also become more concentrated as a function of age (@fig-spatial-parameters).

Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish (@fig-glm-coefficients). When comparing our initial Poisson GLM to a more complicated GLMM with random year effects in the intercept and slope, we found little support for shared correlations across age classes of a given species. Shared correlations would indicate for example that age-2 and age-3 sablefish may be more or less predictable in certain years (indicating that information about one age class may be useful in predicting another). With the exception of a strong positive correlation between age-0 and age-1 sablefish (\$\$)

@@

\newpage

```{r label="fig-hake-spatial-anomaly", fig.cap="Estimated spatial anomalies for Pacific hake ages 1 -- 5. These fields represent the average anomalies across all years, 2003 -- 2023.", echo=FALSE}
knitr::include_graphics("plots/hake_spatial_anomalies.png")
```

\newpage

```{r label="fig-sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies for sablefish ages 0 -- 9. These fields represent the average anomalies across all years, 2003 -- 2023.", echo=FALSE}
knitr::include_graphics("plots/sablefish_spatial_anomalies.png")
```

\newpage

```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial composition for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014) and columns represent predictions for ages 1 - 5. Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
knitr::include_graphics("plots/spatial_composition_subset_Pacific hake.png")
```

\newpage

```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial composition for sablefish; rows represent strong cohorts in our dataset (2013, 2016, 2019) and columns represent predictions for ages 0 - 4. Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
knitr::include_graphics("plots/spatial_composition_subset_sablefish.png")
```

\newpage

@@

\clearpage

\newpage

@@

\newpage

@@

\newpage

```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial composition for Pacific hake; rows represent years and columns represent predictions for ages 1 - 5.", echo=FALSE}
knitr::include_graphics("plots/spatial_composition_Pacific hake.png")
```

\newpage

```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial composition for sablefish; rows represent years and columns represent predictions for ages 0 - 4.", echo=FALSE}

knitr::include_graphics("plots/spatial_composition_sablefish.png")
```
